Seabird assemblages are linked to the major western boundary current off eastern Australia

### 2023-10-17 ms_preprint/ms_preprint.Rmd
ms tidy-up and supplementary material in rmd

---
title: Seabird assemblages are linked to the major western boundary current off eastern Australia
authors:
  - name: Nicholas W. Daudt
    department: Department of Marine Science
    affiliation: University of Otago
    location: Dunedin, Aotearoa New Zealand
    email: nicholaswdaudt@gmail.com
  - name: Eric J. Woehler
    department: Australasian Seabird Group
    affiliation: BirdLife Australia
    location: Hobart, Australia
  - name: Matthew R. Schofield
    department: Department of Mathematics and Statistics
    affiliation: University of Otago
    location: Dunedin, Aotearoa New Zealand
  - name: Robert O. Smith
    department: Department of Marine Science
    affiliation: University of Otago
    location: Dunedin, Aotearoa New Zealand
  - name: Leandro Bugoni
    department: Institute of Biological Sciences
    affiliation: Universidade Federal do Rio Grande--FURG
    location: Rio Grande, RS, Brazil
  - name: William J. Rayment
    department: Department of Marine Science
    affiliation: University of Otago
    location: Dunedin, Aotearoa New Zealand
abstract: |
  **Aim**: To identify assemblages of seabird species and relate them to environmental conditions off eastern Australia, a world seabird hotspot.
  **Methods**: We used 15 seabird at-sea ship-based survey data collected over 37° of latitude, from 2016 to 2021, and fitted Region of Common Profile (RCP) mixture models for each season and type of data (presence-absence and abundance). RCP groups are defined as a set of regions where the probability of encountering a specific species profile is constant within regions, but different amongst them. These groups also vary according to covariates, which in our case included oceanographic, climatic, and physiographic parameters.
  **Results**: Based on 142,646 seabirds recorded, from 80 species, all models suggested two macro-scale assemblages (‘northern’ and ‘southern’), except for the autumn presence-absence model that identified three groups. The model results consistently show a biogeographic transition at ~34°S, near the latitude at which the East Australian Current separates from the Australian continental slope. Sea surface temperatures or sea surface salinities were selected in all final models, further indicating a close relationship between seabird assemblages and water masses. Results from both data types, presence-absence and abundance, resulted in similar spatial and species profile patterns.
  **Conclusion**: RCP models clearly identified two seabird assemblages off the east coast of Australia, suggesting the persistence of these groups at seasonal and macro spatial scales. Given the rapid climatic changes that the East Australian Current is experiencing and its importance in influencing distributions of seabirds, the methods applied in our study could be replicated to assess possible changes in seabird assemblages, and how they are affected by changing environmental conditions.
keywords:
  - Assemblages
  - Community ecology
  - East Australian Current
  - Mixture models
  - Region of Common Profiles
bibliography: ../references.bib
# biblio-style: 
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/apa.csl
output: rticles::arxiv_article
header-includes:
- \usepackage{setspace}\doublespacing
- \usepackage{pdflscape}
- \DeclareUnicodeCharacter{0394}{\ensuremath{\Delta}}
- \newcommand{\blandscape}{\begin{landscape}}
- \newcommand{\elandscape}{\end{landscape}}
---

<!-- Running title: Seabird assemblages off East Australia -->
<!-- Note: I've changed the 'arxiv.sty' source file to include the running title in the PDF -->

@@

The EAC originates from the South Equatorial Current (SEC), a tropical nutrient-poor current. When the SEC's North Caledonian Jet meets the Queensland Plateau (\~18°S), it splits into a north-flowing branch towards the Gulf of Papua and a south-flowing branch, which marks the start of the EAC [@kessler2013; @ganachaud2014]. The South Caledonian Jet completes the EAC system at \~22°S after encountering the shelf break in the southern portion of the Great Barrier Reef (**Fig. 1a**; @kessler2013; @ganachaud2014). **SST/SSS characteristics? ROB?** The EAC interacts with the bathymetry of the seafloor, giving rise to oceanographic features and processes, such as smaller currents, upwellings, and persistent dynamic eddies [@oke2000; @ridgway2003a; @oke2019]. The EAC flow polewards and is primarily associated with the shelf break, but at \~32°S it detaches from the shelf and spawns an area of eddy generation that dissipates throughout the western Tasman Sea [@mata2006; @oke2019], the 'EAC eastern extension' (@oke2019a; often termed the 'Tasman Front'). Although most EAC energy is converted into eddies, the EAC still runs southwards following the shelf break towards Tasmania (the 'EAC southern extension'; **Fig. 1a**), carrying eddy-like features as a transient current [@oke2019].

@@

Seabird data were gathered aboard *RV Investigator* and *RV Falkor* on 15 voyages, between 2016/17--2020/21, spanning over \~37° of latitude (**Fig. 1b**; **Table S1**). One or more observers continuously recorded seabirds within a 300 m forward quadrant (eye height 25 m a.s.l) from sunrise to sunset hours while the ship was underway. The species, the numbers of birds, and their behaviours were logged in real time into a ship-based web server. **Environmental parameters for atmospheric and oceanographic conditions were logged continuously by automated-ship based sensors, with seabird observations supplementing relevant environmental conditions such as visibility or associations with marine mammals.** Individual records received a date/time/geographic coordinate label (multi-species records have the same label but are logged as separate entries for each species). Birds following the vessels were excluded from analyses. Seabird observation protocols followed @raymond2010, in use around Australia since 1980.

To investigate and predict species' distributions and relationships with environmental variables, a considered choice of **broad-scale climatic and oceanographic** covariates is important [@barbet-massin2014]. We used three physiographic variables and seven oceanographic variables, from which a further three climatic layers were calculated (**Table 1**). These variables are the most frequently used in seabird-environment relationship studies [@tremblay2009], capturing processes at a range of spatial and temporal scales from macro-scale water-mass properties to prey availability in meso- to submesoscale oceanographic features [@haury1978patterns; @weimerskirch2007; @scales2014].

@@

Data wrangling, modelling and visualization were done in R 4.2.0 [@r2022]; the full list of packages can be found in the Supplementary Material **XX**. Code and the modelling dataset (see \ref{sec:modellingdatasets}) are available in **Daudt et al. 2023 (Zenodo repo)**.

@@

The spatial and temporal persistence of oceanographic features such as eddy kinetic energy ('eke') and thermal fronts (sea surface temperature gradient: 'sst_grad') may influence the movements and distributions of seabirds by offering a higher likelihood of finding food [@weimerskirch2007; @scales2014a]. Thus, 'eke' and 'sst_grad' environmental layers were calculated before extracting environmental data for each seabird record. For both 'eke' and 'sst_grad', we used their original cell spatial resolution (**Table 1**) and calculations were done layer-by-layer to match seabird data on the finest spatial scale possible. Eddy kinetic energy was calculated as $0.5 (u'² + v'²)$, where $u' = u - \bar{u}$ and $v' = v - \bar{v}$ are the surface zonal (latitudinal) and meridian (longitudinal) geostrophic currents, respectively, using a rolling mean of 90 days as $\bar{u}$ and $\bar{v}$ [@sun2022]. The spatial 'sst_grad' for each cell was calculated by averaging the maximum value for each neighbouring cell (average neighbourhood maximum), from north-to-south (NS) and west-to-east (WE); then, the magnitude of the gradient was defined as $NS² + WE²$ [@burrows2011]. In addition, we calculated climatic seasonal average layers for 'eke' and 'sst_grad' based on the 11-year data for 2011--2021 as a proxy for persistent oceanographic features, and the 'eke' standard deviation over the same period (named with the prefix 'clim\_' in **Table 1**). Seabed slope was calculated using 'raster::terrain()' [@raster], and distance from the coast was calculated through 'sf::st_distance()' [@sf] using Australia and Tasmania as the mainland (i.e. ignoring smaller islands). Then, for each seabird record, the most proximate value for each environmental layer (original or calculated) was extracted using built-in functions from 'raster' 3.5-21, 'terra' 1.6-7, and 'sf' 1.0-8 packages [@sf; @raster; @terra].

### Data {#modellingdatasets}
\label{sec:modellingdatasets}

Seabirds are highly mobile animals, and many species migrate and/or disperse during their non-breeding periods [@biology2002]. Therefore, seasonal differences in species composition were expected [e.g. @reid2023], and in the probabilities of occurrence and in total numbers. Thus, we built separate modelling dataset for each season. We aggregated seabird data by 1° latitude $\times$ 1° longitude grid cells (c. 110 $\times$ 110 km) and season (**Fig. 1b**), by summing seabirds from the same species. To test if the presence-absence and abundance (raw counts) data would result in similar assemblage patterns, we built two species matrices, where the presence-absence matrix transformed values \>1 to 1. To avoid numerical instabilities in the models prompted by unusual or rare species, we kept only species occurring in at least six grids within each season (summer = 12 species; autumn = 21; winter = 20; spring = 33). The associated environmental data of each seabird record were averaged within the same grid cells and seasons, and environmental data were standardised before modelling.

@@

A total of 10,261 seabird records was obtained during the 15 voyages, comprising 142,646 individuals from 80 species, belonging to 11 families and five orders. The five most abundant species made up 83% of total birds counted (43.0% short-tailed shearwater *Ardenna tenuirostris*, 17.2% fairy prion *Pachyptila turtur*, 14.3% wedge-tailed shearwater *Ardenna pacifica*, 5.2% sooty tern *Onychoprion fuscatus*, 3.2% shy albatross *Thalassarche cauta*). The five most frequently encountered species were *A. pacifica* (frequency of occurrence = 35.3%), *T. cauta* (33.2%), brown booby *Sula leucogaster* (25.4%), red-footed booby *S. sula* (25.4%) and black-browed albatross *T. melanophris* (25%) (**Fig. S1**). An increase in species richness towards the south of the study region was clear in all but the winter season (**Fig S2**). Noticeably higher counts were made around Tasmania, possibly due to the 'colony effect' and/or the presence of Southern Ocean species (**Fig. S2**). However, one cell in the Coral Sea had counts of the same order of magnitude as off Tasmania in summer.

@@

For all models, two RCPs was the optimal number of seabird groups, except for the presence-absence-autumn model, which indicated three RCPs as the best fit (**Fig. S3**). The best fitting models generally had 1 or 2 covariates selected, although the abundance-autumn model had five covariates (**Table 2**). The best models showed a good fit with no apparent deviations from the assumptions (**Fig. S4**). All the best models selected 'sst' or 'sss', as essential for characterising water masses. The probability of belonging to one or another RCP varied its 'sst' threshold between 18--21 °C, or 'sss' 35.1 PSU for summer (**Fig. 2**, **Fig. S5**), similar to the EAC's temperature and salinity thresholds. Furthermore, most of the best models selected 'bat', suggesting a gradient between 'coastal' and 'oceanic' taxa based on bathymetry (**Fig. S5**). Of particular note, proxies for persistent oceanographic features that may aggregate food resources (the 'climatic' variables) for seabirds were selected only in the autumn and spring abundance models.

Grid point-predictions are the assignment of the RCP with the highest probability value for a given cell. When predicting RCP onto space, patterns of point-predictions suggest a biogeographic transition in seabird species around \~34°S, with both presence-absence and abundance models pointing to similar results (**Fig. 3**). These patterns are further reinforced when looking at the mean probability along with the lower and upper confidence intervals of predictions (**Fig. 4**, **Fig. S6**), showing that around \~34°S, the 'northern' RCP tended to have higher probabilities in coastal regions, whereas the 'southern' RCP spread northward offshore. The presence-absence-autumn model (three RCPs) had a similar broad-scale pattern, although the southern RCP is farther south, and a strong 'Tasmania/Bass Straight' assemblage appears (**Fig. 3**, **Fig. S6**).

The broad pattern of two large-scale assemblages is reflected in their species compositions. Although there are overlaps among the seabird groups, both presence-absence and abundance models captured typical tropical species in the northern RCP (e.g. boobies *Sula* spp., lesser frigatebird *Fregata ariel*) and typical Southern Ocean species in the southern RCP (e.g. small albatrosses *Thalassarche* spp., fairy prion *P. turtur*) (**Fig. 5**, **Fig. S7**). The presence-absence-autumn model, which had three RCPs, also showed the characteristic taxa representing the 'northern' and 'southern' RCPs. However, the intermediate 'Tasmania/Bass Straight' assemblage showed a mixture of both groups with shy albatross and Australasian gannet *Morus serrator* prominent with high probabilities of occurrence.

@@

## Limitations
\label{sec:limitations}

@@

<div id="refs"></div>

@@

# Tables and Figures {-}

@@

data.frame(
  Variable = c("Sea Surface Temperature", "Sea Surface Temperature gradient", "Sea Surface Salinity", "Mixed Layer Depth", "8-day average Chlorophyll-a", "Log10 8-day average Chlorophyll-a", "Eddy Kinetic Energy", "Climatic mean EKE", "Climatic standard deviation EKE", "Climatic mean SSTgrad", "Bathymetry", "Slope", "Distance from coast"),
  # Abbreviation needs the `str_replace_all` because I'm using `escape=F` in `kbl()`
  Abbreviation = stringr::str_replace_all(c("sst", "sst_grad", "sss", "mld", "chl8", "log10_chl8", "eke", "clim_eke_mean", "clim_eke_sd", "clim_sst_grad", "bat", "slope", "dist_coast"), "_", "\\\\_"), 
  Units = c("°Celsius", paste0(parse(text = "\u0394"), "°Celsius"), "PSU", "m", "mg/m${^3}$", "$log_{10}$(chl8)", "m/s", "mean(m/s)/season/10 years", "sd(m/s)/season/10 years", paste0("mean(", paste0(parse(text = "\u0394"), "°Celsius"), ")/season/10 years"), "m", "degrees (°)", "km"),
  `Spatial resolution` = c("0.25° × 0.25°", "0.25° × 0.25°", "0.083° × 0.083°", "0.083° × 0.083°", "0.036° × 0.036°", "0.036° × 0.036°", "0.2° × 0.2°", "0.2° × 0.2°", "0.2° × 0.2°", "0.25° × 0.25°", "0.004° × 0.004°", "0.004° × 0.004°", "NA"),
  `Temporal resolution` = c("daily", "daily", "daily", "daily", "8-days composite", "8-days composite", "daily", "seasonal climatology", "seasonal climatology", "seasonal climatology", "NA", "NA", "NA"),
  Source = c("Huang et al. (2021)${^a}$", "Derived from Huang et al. (2021)${^a}$", "E.U. Copernicus Marine Service Information${^b}{^c}$", "E.U. Copernicus Marine Service Information${^b}{^c}$", "ERDDAP/NOAA${^d}$", "Derived from ERDDAP/NOAA${^d}$", "IMOS/AODN${^e}$", "Derived from IMOS/AODN${^e}$", "Derived from IMOS/AODN${^e}$", "Derived from Huang et al. (2021)${^a}$", "GEBCO${^f}$", "Derived from GEBCO${^f}$", "Calculated in R (see main text)")
) |>
  kableExtra::kbl(caption = " ", booktabs = TRUE, escape = FALSE)  %>% 
  kableExtra::row_spec(0, bold = TRUE) %>% 
  kableExtra::column_spec(1, width = "4cm") %>% 
  kableExtra::column_spec(6, width = "3cm") %>% 
  kableExtra::footnote(
    alphabet = c(
    "NOAA OISST: (https://psl.noaa.gov/data/gridded/data.noaa.oisst.v2.highres.html); ", 
    "2011--2019: GLOBAL_MULTIYEAR_PHY_001_030 (https://doi.org/10.48670/moi-00021); ", 
    "2020--2021: GLOBAL_ANALYSIS_FORECAST_PHY_001_024 (https://doi.org/10.48670/moi-00016); ",
    "ERDDAP/NOAA (https://coastwatch.pfeg.noaa.gov/erddap/griddap/erdMH1chla8day.html); ",
    "IMOS/AODN (https://portal.aodn.org.au/search), 'OceanCurrent - Gridded sea level anomaly - Delayed mode' data set; ",
    "GEBCO (https://www.gebco.net/data_and_products/gridded_bathymetry_data/)"))

>>>>

---
title: Seabird assemblages are linked to the major western boundary current off eastern Australia
authors:
  - name: Nicholas W. Daudt
    department: Department of Marine Science
    affiliation: University of Otago
    location: Dunedin, Aotearoa New Zealand
    email: nicholaswdaudt@gmail.com
  - name: Eric J. Woehler
    department: Australasian Seabird Group
    affiliation: BirdLife Australia
    location: Hobart, Australia
  - name: Matthew R. Schofield
    department: Department of Mathematics and Statistics
    affiliation: University of Otago
    location: Dunedin, Aotearoa New Zealand
  - name: Robert O. Smith
    department: Department of Marine Science
    affiliation: University of Otago
    location: Dunedin, Aotearoa New Zealand
  - name: Leandro Bugoni
    department: Institute of Biological Sciences
    affiliation: Universidade Federal do Rio Grande--FURG
    location: Rio Grande, RS, Brazil
  - name: William J. Rayment
    department: Department of Marine Science
    affiliation: University of Otago
    location: Dunedin, Aotearoa New Zealand
abstract: |
  **Aim**: To identify assemblages of seabird species and relate them to environmental conditions off eastern Australia, a world seabird hotspot.
  **Methods**: We used 15 seabird at-sea ship-based survey data collected over 37° of latitude, from 2016 to 2021, and fitted Region of Common Profile (RCP) mixture models for each season and type of data (presence-absence and abundance). RCP groups are defined as a set of regions where the probability of encountering a specific species profile is constant within regions, but different amongst them. These groups also vary according to covariates, which in our case included oceanographic, climatic, and physiographic parameters.
  **Results**: Based on 142,646 seabirds recorded, from 80 species, all models suggested two macro-scale assemblages (‘northern’ and ‘southern’), except for the autumn presence-absence model that identified three groups. The model results consistently show a biogeographic transition at ~34°S, near the latitude at which the East Australian Current separates from the Australian continental slope. Sea surface temperatures or sea surface salinities were selected in all final models, further indicating a close relationship between seabird assemblages and water masses. Results from both data types, presence-absence and abundance, resulted in similar spatial and species profile patterns.
  **Conclusion**: RCP models clearly identified two seabird assemblages off the east coast of Australia, suggesting the persistence of these groups at seasonal and macro spatial scales. Given the rapid climatic changes that the East Australian Current is experiencing and its importance in influencing distributions of seabirds, the methods applied in our study could be replicated to assess possible changes in seabird assemblages, and how they are affected by changing environmental conditions.
keywords:
  - Assemblages
  - Biogeography
  - Community ecology
  - East Australian Current
  - Mixture models
  - Region of Common Profiles
bibliography: ../references.bib
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/apa.csl
# output: rticles::arxiv_article
output:
  bookdown::pdf_book:
    base_format: rticles::arxiv_article
header-includes:
- \usepackage{setspace}\doublespacing
- \usepackage{pdflscape}
- \DeclareUnicodeCharacter{0394}{\ensuremath{\Delta}} # needed to understand unicode for greek Delta in Table 1
- \newcommand{\blandscape}{\begin{landscape}}
- \newcommand{\elandscape}{\end{landscape}}
---

<!-- Running title: Seabird assemblages off Eastern Australia -->

<!-- Note: I've changed the 'arxiv.sty' source file to include the running title in the PDF -->

@@

The EAC originates from the South Equatorial Current (SEC), a tropical nutrient-poor current. When the SEC's North Caledonian Jet meets the Queensland Plateau (\~18°S), it splits into a north-flowing branch towards the Gulf of Papua and a south-flowing branch, which marks the start of the EAC [@kessler2013; @ganachaud2014]. The South Caledonian Jet completes the EAC system at \~22°S after encountering the shelf break in the southern portion of the Great Barrier Reef (**Fig. \@ref(fig:fig1-study-area)a**; @kessler2013; @ganachaud2014). **SST/SSS characteristics? ROB?** The EAC interacts with the bathymetry of the seafloor, giving rise to oceanographic features and processes, such as smaller currents, upwellings, and persistent dynamic eddies [@oke2000; @ridgway2003a; @oke2019]. The EAC flow polewards and is primarily associated with the shelf break, but at \~32°S it detaches from the shelf and spawns an area of eddy generation that dissipates throughout the western Tasman Sea [@mata2006; @oke2019], the 'EAC eastern extension' (@oke2019a; often termed the 'Tasman Front'). Although most EAC energy is converted into eddies, the EAC still runs southwards following the shelf break towards Tasmania (the 'EAC southern extension'; **Fig. \@ref(fig:fig1-study-area)a**), carrying eddy-like features as a transient current [@oke2019].

@@

Seabird data were gathered aboard *RV Investigator* and *RV Falkor* on 15 voyages, between 2016/17--2020/21, spanning over \~37° of latitude (**Fig. \@ref(fig:fig1-study-area)b**; **Table S1**). One or more observers continuously recorded seabirds within a 300 m forward quadrant (eye height 25 m a.s.l) from sunrise to sunset hours while the ship was underway. The species, the numbers of birds, and their behaviours were logged in real time into a ship-based web server. Environmental parameters for atmospheric and oceanographic conditions were logged continuously by automated-ship based sensors, with seabird observations supplementing relevant environmental conditions such as visibility or associations with marine mammals. Individual records received a date/time/geographic coordinate label (multi-species records have the same label but are logged as separate entries for each species). Birds following the vessels were excluded from analyses. Seabird observation protocols followed @raymond2010, in use around Australia since 1980.

To investigate and predict species' distributions and relationships with environmental variables, a considered choice of broad-scale climatic and oceanographic covariates is important [@barbet-massin2014]. We used three physiographic variables and seven oceanographic variables, from which a further three climatic layers were calculated (**Table \@ref(tab:table1)**). These variables are the most frequently used in seabird-environment relationship studies [@tremblay2009], capturing processes at a range of spatial and temporal scales from macro-scale water-mass properties to prey availability in meso- to submesoscale oceanographic features [@haury1978patterns; @weimerskirch2007; @scales2014].

@@

Data wrangling, modelling and visualization were done in R 4.2.0 [@r2022]; the full list of packages can be found in the Supplementary Material. Code and the modelling dataset (see \ref{sec:modellingdatasets}) are available in **(Zenodo repo)**.

@@

The spatial and temporal persistence of oceanographic features such as eddy kinetic energy ('eke') and thermal fronts (sea surface temperature gradient: 'sst_grad') may influence the movements and distributions of seabirds by offering a higher likelihood of finding food [@weimerskirch2007; @scales2014a]. Thus, 'eke' and 'sst_grad' environmental layers were calculated before extracting environmental data for each seabird record. For both 'eke' and 'sst_grad', we used their original cell spatial resolution (**Table \@ref(tab:table1)**) and calculations were done layer-by-layer to match seabird data on the finest spatial scale possible. Eddy kinetic energy was calculated as $0.5 (u'² + v'²)$, where $u' = u - \bar{u}$ and $v' = v - \bar{v}$ are the surface zonal (latitudinal) and meridian (longitudinal) geostrophic currents, respectively, using a rolling mean of 90 days as $\bar{u}$ and $\bar{v}$ [@sun2022]. The spatial 'sst_grad' for each cell was calculated by averaging the maximum value for each neighbouring cell (average neighbourhood maximum), from north-to-south (NS) and west-to-east (WE); then, the magnitude of the gradient was defined as $NS² + WE²$ [@burrows2011]. In addition, we calculated climatic seasonal average layers for 'eke' and 'sst_grad' based on the 11-year data for 2011--2021 as a proxy for persistent oceanographic features, and the 'eke' standard deviation over the same period (named with the prefix 'clim\_' in **Table \@ref(tab:table1)**). Seabed slope was calculated using 'raster::terrain()' [@raster], and distance from the coast was calculated through 'sf::st_distance()' [@sf] using Australia and Tasmania as the mainland (i.e. ignoring smaller islands). Then, for each seabird record, the most proximate value for each environmental layer (original or calculated) was extracted using built-in functions from 'raster' 3.5-21, 'terra' 1.6-7, and 'sf' 1.0-8 packages [@sf; @raster; @terra].

### Data {#modellingdatasets}

\label{sec:modellingdatasets}

Seabirds are highly mobile animals, and many species migrate and/or disperse during their non-breeding periods [@biology2002]. Therefore, seasonal differences in species composition were expected [e.g. @reid2023], and in the probabilities of occurrence and in total numbers. Thus, we built separate modelling dataset for each season. We aggregated seabird data by 1° latitude $\times$ 1° longitude grid cells (c. 110 $\times$ 110 km) and season (**Fig. \@ref(fig:fig1-study-area)b**), by summing seabirds from the same species. To test if the presence-absence and abundance (raw counts) data would result in similar assemblage patterns, we built two species matrices, where the presence-absence matrix transformed values \>1 to 1. To avoid numerical instabilities in the models prompted by unusual or rare species, we kept only species occurring in at least six grids within each season (summer = 12 species; autumn = 21; winter = 20; spring = 33). The associated environmental data of each seabird record were averaged within the same grid cells and seasons, and environmental data were standardised before modelling.

@@

A total of 10,261 seabird records was obtained during the 15 voyages, comprising 142,646 individuals from 80 species, belonging to 11 families and five orders. The five most abundant species made up 83% of total birds counted (43.0% short-tailed shearwater *Ardenna tenuirostris*, 17.2% fairy prion *Pachyptila turtur*, 14.3% wedge-tailed shearwater *Ardenna pacifica*, 5.2% sooty tern *Onychoprion fuscatus*, 3.2% shy albatross *Thalassarche cauta*). The five most frequently encountered species were *A. pacifica* (frequency of occurrence = 35.3%), *T. cauta* (33.2%), brown booby *Sula leucogaster* (25.4%), red-footed booby *S. sula* (25.4%) and black-browed albatross *T. melanophris* (25%) (**Fig. S1**). An increase in species richness towards the south of the study region was clear in all but the winter season (**Fig. S2**). Noticeably higher counts were made around Tasmania, possibly due to the 'colony effect' and/or the presence of Southern Ocean species (**Fig. S2**). However, one cell in the Coral Sea had counts of the same order of magnitude as off Tasmania in summer.

@@

For all models, two RCPs was the optimal number of seabird groups, except for the presence-absence-autumn model, which indicated three RCPs as the best fit (**Fig. S3**). The best fitting models generally had 1 or 2 covariates selected, although the abundance-autumn model had five covariates (**Table \@ref(tab:table2)**). The best models showed a good fit with no apparent deviations from the assumptions (**Fig. S4**). All the best models selected 'sst' or 'sss', as essential for characterising water masses. The probability of belonging to one or another RCP varied its 'sst' threshold between 18--21 °C, or 'sss' 35.1 PSU for summer (**Fig. \@ref(fig:fig2-partial-plots)**, **Fig. S5**), similar to the EAC's temperature and salinity thresholds. Furthermore, most of the best models selected 'bat', suggesting a gradient between 'coastal' and 'oceanic' taxa based on bathymetry (**Fig. S5**). Of particular note, proxies for persistent oceanographic features that may aggregate food resources (the 'climatic' variables) for seabirds were selected only in the autumn and spring abundance models.

Grid point-predictions are the assignment of the RCP with the highest probability value for a given cell. When predicting RCP onto space, patterns of point-predictions suggest a biogeographic transition in seabird species around \~34°S, with both presence-absence and abundance models pointing to similar results (**Fig. \@ref(fig:fig3-point-pred)**). These patterns are further reinforced when looking at the mean probability along with the lower and upper confidence intervals of predictions (**Fig. \@ref(fig:fig4-prob-maps)**, **Fig. S6**), showing that around \~34°S, the 'northern' RCP tended to have higher probabilities in coastal regions, whereas the 'southern' RCP spread northward offshore. The presence-absence-autumn model (three RCPs) had a similar broad-scale pattern, although the southern RCP is farther south, and a strong 'Tasmania/Bass Straight' assemblage appears (**Fig. \@ref(fig:fig3-point-pred)**, **Fig. S6**).

The broad pattern of two large-scale assemblages is reflected in their species compositions. Although there are overlaps among the seabird groups, both presence-absence and abundance models captured typical tropical species in the northern RCP (e.g. boobies *Sula* spp., lesser frigatebird *Fregata ariel*) and typical Southern Ocean species in the southern RCP (e.g. small albatrosses *Thalassarche* spp., fairy prion *P. turtur*) (**Fig. \@ref(fig:fig5-spp-profiles)**, **Fig. S7**). The presence-absence-autumn model, which had three RCPs, also showed the characteristic taxa representing the 'northern' and 'southern' RCPs. However, the intermediate 'Tasmania/Bass Straight' assemblage showed a mixture of both groups with shy albatross and Australasian gannet *Morus serrator* prominent with high probabilities of occurrence.

@@

## Limitations

\label{sec:limitations}

@@

::: {#refs}
:::

@@

# Tables and Figures {.unnumbered}

@@

data.frame(
  Variable = c("Sea Surface Temperature", "Sea Surface Temperature gradient", "Sea Surface Salinity", "Mixed Layer Depth", "8-day average Chlorophyll-a", "Log10 8-day average Chlorophyll-a", "Eddy Kinetic Energy", "Climatic mean EKE", "Climatic standard deviation EKE", "Climatic mean SSTgrad", "Bathymetry", "Slope", "Distance from coast"),
  # Abbreviation needs the `str_replace_all` because I'm using `escape=F` in `kbl()`
  Abbreviation = stringr::str_replace_all(c("sst", "sst_grad", "sss", "mld", "chl8", "log10_chl8", "eke", "clim_eke_mean", "clim_eke_sd", "clim_sst_grad", "bat", "slope", "dist_coast"), "_", "\\\\_"), 
  Units = c("°Celsius", paste0(parse(text = "\u0394"), "°Celsius"), "PSU", "m", "mg/m${^3}$", "$log_{10}$(chl8)", "m/s", "mean(m/s)/season/10 years", "sd(m/s)/season/10 years", paste0("mean(", paste0(parse(text = "\u0394"), "°Celsius"), ")/season/10 years"), "m", "degrees (°)", "km"),
  `Spatial resolution` = c("0.25° × 0.25°", "0.25° × 0.25°", "0.083° × 0.083°", "0.083° × 0.083°", "0.036° × 0.036°", "0.036° × 0.036°", "0.2° × 0.2°", "0.2° × 0.2°", "0.2° × 0.2°", "0.25° × 0.25°", "0.004° × 0.004°", "0.004° × 0.004°", "NA"),
  `Temporal resolution` = c("daily", "daily", "daily", "daily", "8-days composite", "8-days composite", "daily", "seasonal climatology", "seasonal climatology", "seasonal climatology", "NA", "NA", "NA"),
  Source = c("Huang et al. (2021)${^a}$", "Derived from Huang et al. (2021)${^a}$", "E.U. Copernicus Marine Service Information${^b}{^c}$", "E.U. Copernicus Marine Service Information${^b}{^c}$", "ERDDAP/NOAA${^d}$", "Derived from ERDDAP/NOAA${^d}$", "IMOS/AODN${^e}$", "Derived from IMOS/AODN${^e}$", "Derived from IMOS/AODN${^e}$", "Derived from Huang et al. (2021)${^a}$", "GEBCO${^f}$", "Derived from GEBCO${^f}$", "Calculated in R (see main text)")
) %>% 
  kableExtra::kbl(caption = " ", booktabs = TRUE, escape = FALSE)  %>% 
  kableExtra::row_spec(0, bold = TRUE) %>% 
  kableExtra::column_spec(1, width = "4cm") %>% 
  kableExtra::column_spec(6, width = "3cm") %>% 
  kableExtra::footnote(
    alphabet = c(
    "NOAA OISST: (https://psl.noaa.gov/data/gridded/data.noaa.oisst.v2.highres.html); ", 
    "2011--2019: GLOBAL_MULTIYEAR_PHY_001_030 (https://doi.org/10.48670/moi-00021); ", 
    "2020--2021: GLOBAL_ANALYSIS_FORECAST_PHY_001_024 (https://doi.org/10.48670/moi-00016); ",
    "ERDDAP/NOAA (https://coastwatch.pfeg.noaa.gov/erddap/griddap/erdMH1chla8day.html); ",
    "IMOS/AODN (https://portal.aodn.org.au/search), 'OceanCurrent - Gridded sea level anomaly - Delayed mode' data set; ",
    "GEBCO (https://www.gebco.net/data_and_products/gridded_bathymetry_data/)"))
